---
title: "Brainstorm 2: Is GitHub's tedious, high-demand work outsourced to the Global South?"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    df-print: paged
    fig-width: 10
    fig-height: 6
editor:
  markdown:
    wrap: 80
---

## The hypothesis

> *High-demand, tedious tech work is mostly outsourced to the Global South.*

If true, GitHub activity in Global-South economies should disproportionately
sit on **maintenance-coded, body-shop stacks** — PHP (WordPress / Drupal /
Magento), legacy Java, HTML/CSS templating — while Global-North activity
should over-index on **R&D / cutting-edge stacks** — Rust, Go, Swift, modern
ML/LLM topics.

We test this with four reads of the data:

1.  **Maps** — one panel per UNDP region supercluster, each in its own
    region-appropriate projection, coloring each country by its
    *maintenance-stack share*.
2.  **Head-to-head bars** — North vs. South on each of the suspect
    languages, in raw percentages.
3.  **Country scatter** — every country plotted on *(elite-stack share,
    maintenance-stack share)*.
4.  **Per-topic South-lift** — which specific topics over-index in the
    South vs. the North.

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(scales)
library(countrycode)
library(sf)
library(rnaturalearth)
library(patchwork)
library(ggrepel)
library(ggiraph)

theme_set(theme_minimal(base_size = 11))
```

## Data + Global N/S definition

We use UNDP's own classification: in `hdi.csv` the 6 developing-region
codes (SSA, LAC, AS, SA, EAP, ECA-developing) are the **Global South**;
countries with `region = NA` (Europe, North America, Japan, Korea,
Australia/NZ, Israel, etc. — the developed set) are the **Global North**.
This is a defensible, source-of-record split.

```{r data, message=FALSE, warning=FALSE}
langs   <- read_csv("languages.csv",   show_col_types = FALSE) |>
  filter(iso2_code != "EU")
topics  <- read_csv("topics.csv",      show_col_types = FALSE) |>
  filter(iso2_code != "EU", !topic %in% c("config", "github-config"))
devs    <- read_csv("developers.csv",  show_col_types = FALSE) |>
  filter(iso2_code != "EU")
hdi_raw <- read_csv("hdi.csv",         show_col_types = FALSE)

hdi <- hdi_raw |>
  filter(!is.na(iso3), nchar(iso3) == 3) |>
  transmute(
    iso2_code = countrycode(iso3, "iso3c", "iso2c"),
    country, region_undp = region,
    hemisphere = if_else(is.na(region), "Global North", "Global South"),
    hdi_tier   = factor(hdicode, levels = c("Low", "Medium", "High", "Very High")),
    hdi        = hdi_2022
  ) |>
  filter(!is.na(iso2_code), !is.na(hdi))

latest <- langs |> slice_max(year * 10 + quarter, n = 1) |>
  distinct(year, quarter)

langs_snap  <- langs  |> semi_join(latest, by = c("year", "quarter")) |>
  inner_join(hdi, by = "iso2_code")
topics_snap <- topics |> semi_join(latest, by = c("year", "quarter")) |>
  inner_join(hdi, by = "iso2_code")
devs_snap   <- devs   |> semi_join(latest, by = c("year", "quarter")) |>
  inner_join(hdi, by = "iso2_code")
```

How big is each side?

```{r}
devs_snap |>
  group_by(hemisphere) |>
  summarise(countries  = n(),
            developers = comma(sum(developers)),
            .groups    = "drop")
```

## What counts as "maintenance" vs. "elite"

These are explicit, defensible buckets. The point isn't that PHP is bad or
Rust is good — it's that the *commercial demand profile* differs:

-   **Maintenance / high-demand stacks** — kept in production for decades
    by armies of contractors, dominated by CMS work and legacy enterprise:
    -   **PHP** (WordPress runs ~40% of the web; Drupal, Magento, Laravel)
    -   **Java** (enterprise Spring apps, Android maintenance)
    -   **HTML, CSS** (theming, templating, slicing designs)
-   **Elite / R&D-coded stacks** — newer, narrower hiring market, used at
    modern infra companies and research labs:
    -   **Rust** (systems R&D, modern infrastructure)
    -   **Go** (cloud-native, Kubernetes / Docker / Terraform)
    -   **Swift** (Apple platform — wealthy-device-market work)
    -   **R** (statistics / research / academia)

```{r}
maint_langs <- c("PHP", "Java", "HTML", "CSS")
elite_langs <- c("Rust", "Go", "Swift", "R")
```

```{r}
share <- langs_snap |>
  group_by(iso2_code, country, hemisphere, hdi, hdi_tier, region_undp) |>
  summarise(total = sum(num_pushers),
            maint = sum(num_pushers[language %in% maint_langs]),
            elite = sum(num_pushers[language %in% elite_langs]),
            .groups = "drop") |>
  mutate(maint_share = maint / total,
         elite_share = elite / total) |>
  filter(total >= 500)
```

------------------------------------------------------------------------

## Plot 1 — Maintenance-stack share, by region (each in its own projection)

Each panel: % of that country's quarterly language-pushers writing
PHP, Java, HTML or CSS. Same color scale across all panels. Each panel
uses a projection chosen for the region (Lambert Azimuthal Equal Area
centered on the region's geographic core; Lambert Conformal Conic for
Europe).

```{r build-map-data, message=FALSE, warning=FALSE}
world <- ne_countries(scale = "medium", returnclass = "sf") |>
  filter(iso_a2 != "-99") |>
  select(iso_a2, name, geometry)

ascii <- function(x) iconv(x, from = "UTF-8", to = "ASCII//TRANSLIT", sub = "?")

map_df <- world |>
  left_join(share |> select(iso_a2 = iso2_code, country, hemisphere,
                            region_undp, maint_share, total),
            by = "iso_a2") |>
  mutate(
    label_country = ascii(coalesce(country, name)),
    tooltip = if_else(is.na(maint_share),
                      paste0(label_country, " - no data"),
                      paste0(label_country, " - ",
                             percent(maint_share, accuracy = 1)))
  )

fill_scale <- scale_fill_gradient2(
  low      = "#1a9850",
  mid      = "#ffffbf",
  high     = "#d73027",
  midpoint = 0.5,
  limits   = c(0, 1),
  oob      = scales::squish,
  labels   = label_percent(accuracy = 1),
  breaks   = c(0, 0.25, 0.5, 0.75, 1),
  name     = "Maintenance-stack\nshare of pushers\n(PHP + Java + HTML + CSS)\n\nGreen < 50% < Red",
  na.value = "grey92"
)

panel <- function(iso_codes, crs, title, show_legend = FALSE) {
  d <- map_df |> filter(iso_a2 %in% iso_codes)
  d_proj <- st_transform(d, crs = crs)
  bb <- st_bbox(d_proj)
  ggplot(d_proj) +
    geom_sf_interactive(
      aes(fill = maint_share, tooltip = tooltip, data_id = iso_a2),
      color = "white", linewidth = 0.15
    ) +
    fill_scale +
    coord_sf(xlim = c(bb["xmin"], bb["xmax"]),
             ylim = c(bb["ymin"], bb["ymax"]),
             expand = FALSE, crs = crs) +
    labs(title = title) +
    theme_void(base_size = 10) +
    theme(plot.title = element_text(face = "bold", size = 10,
                                    margin = margin(b = 4)),
          legend.position = if (show_legend) "right" else "none",
          legend.key.height = unit(1.2, "cm"),
          legend.key.width  = unit(0.4, "cm"),
          legend.title = element_text(size = 8),
          legend.text  = element_text(size = 8),
          panel.background = element_rect(fill = "#f6f7f9", color = NA),
          plot.background  = element_rect(fill = "white", color = NA))
}
```

```{r region-iso-lists}
region_iso <- hdi |>
  group_by(region_undp) |>
  summarise(iso = list(iso2_code), .groups = "drop")

iso_of <- function(r) region_iso |>
  filter(region_undp == r) |> pull(iso) |> unlist()

# Geographic — not political — panel assignments. Russia, Central Asia,
# and the Caucasus go in the Asia panel (so the European projection
# stays readable). Israel goes in the Middle East panel.
iso_na <- c("US", "CA")

iso_europe <- c(
  # Western / Northern / Southern Europe
  "AT","BE","CH","DE","DK","ES","FI","FR","GB","GR","IE","IS",
  "IT","LI","LU","MC","MT","NL","NO","PT","SE","SM","AD","VA",
  "CY",
  # Central / Eastern Europe (incl. Balkans, Baltics, Ukraine, Belarus)
  "AL","BA","BG","BY","CZ","EE","HR","HU","LT","LV","MD","ME",
  "MK","PL","RO","RS","SI","SK","UA","XK",
  # Turkey — straddles, include for east edge
  "TR"
)

iso_asia_pac <- c(
  iso_of("EAP"),
  "JP","KR","HK","TW","SG","AU","NZ",
  "RU",                                # Russia → Asia panel
  "KZ","UZ","KG","TJ","TM","MN",       # Central Asia + Mongolia
  "GE","AM","AZ"                       # Caucasus
)

iso_middle_east <- c(iso_of("AS"), "IL", "IR")
```

```{r fig-maps, fig.width=14, fig.height=11, message=FALSE, warning=FALSE}
p_na   <- panel(iso_na,
                crs = "+proj=laea +lat_0=45 +lon_0=-100",
                title = "North America")

p_eu   <- panel(iso_europe,
                crs = 3035,
                title = "Europe")

p_eap  <- panel(iso_asia_pac,
                crs = "+proj=laea +lat_0=30 +lon_0=110",
                title = "Asia & Pacific (incl. Russia, Central Asia)")

p_lac  <- panel(iso_of("LAC"),
                crs = "+proj=laea +lat_0=-15 +lon_0=-60",
                title = "Latin America & Caribbean")

p_arab <- panel(iso_middle_east,
                crs = "+proj=laea +lat_0=27 +lon_0=45",
                title = "Middle East")

p_sa   <- panel(iso_of("SA"),
                crs = "+proj=laea +lat_0=22 +lon_0=80",
                title = "South Asia")

p_ssa  <- panel(iso_of("SSA"),
                crs = "+proj=laea +lat_0=0 +lon_0=20",
                title = "Sub-Saharan Africa",
                show_legend = TRUE)
```

```{r fig-map-assembled, fig.width=14, fig.height=11}
map_grid <- ((p_na   | p_eu   | p_eap) /
             (p_lac  | p_arab | p_sa)  /
             (p_ssa  | plot_spacer() | plot_spacer())) +
  plot_annotation(
    title    = "Where on Earth is the 'maintenance stack' actually written?",
    subtitle = "Per-country share of pushers on PHP + Java + HTML + CSS — each region in its own projection. Hover for country + share.",
    theme    = theme(plot.title    = element_text(face = "bold", size = 14),
                     plot.subtitle = element_text(size = 10, color = "grey30"))
  )

girafe(
  ggobj = map_grid,
  width_svg = 14, height_svg = 11,
  options = list(
    opts_tooltip(css = "background-color:#1a1a1a;color:white;padding:4px 8px;border-radius:4px;font-family:sans-serif;font-size:12px;"),
    opts_hover(css = "stroke:#111111;stroke-width:1.2px;"),
    opts_zoom(min = 1, max = 4)
  )
)
```

------------------------------------------------------------------------

## Plot 2 — Head-to-head bars on the suspect languages

For each of our 8 suspect languages, what share of pushers in Global North
vs. Global South spend their quarter pushing it? This is the raw "do they
write the same things?" picture.

```{r fig-bars, fig.width=10, fig.height=5}
suspect <- c(maint_langs, elite_langs)

hem_totals_lang <- langs_snap |>
  group_by(hemisphere) |>
  summarise(hem_total = sum(num_pushers), .groups = "drop")

shares <- langs_snap |>
  filter(language %in% suspect) |>
  group_by(hemisphere, language) |>
  summarise(pushers = sum(num_pushers), .groups = "drop") |>
  inner_join(hem_totals_lang, by = "hemisphere") |>
  mutate(share  = pushers / hem_total,
         bucket = factor(if_else(language %in% maint_langs,
                                 "Maintenance-coded", "Elite / R&D-coded"),
                         levels = c("Maintenance-coded", "Elite / R&D-coded")),
         language = factor(language,
                           levels = c(maint_langs, elite_langs)))

ggplot(shares, aes(language, share, fill = hemisphere)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(aes(label = percent(share, accuracy = 0.1)),
            position = position_dodge(width = 0.7),
            vjust = -0.3, size = 3) +
  facet_grid(. ~ bucket, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = c("Global North" = "#2c7fb8",
                               "Global South" = "#e34a33"),
                    name = NULL) +
  scale_y_continuous(labels = label_percent(accuracy = 1),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "Share of pushers on each language: Global North vs. Global South",
    subtitle = "If the hypothesis holds, the South should win the left facet and lose the right",
    x = NULL, y = "Share of hemisphere's pushers"
  ) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))
```

A different framing: the **South / North ratio** for each language. >1
means the language is disproportionately Southern; <1 means Northern.

```{r fig-ratio, fig.width=8, fig.height=4}
shares |>
  select(hemisphere, language, share, bucket) |>
  pivot_wider(names_from = hemisphere, values_from = share) |>
  mutate(ratio = `Global South` / `Global North`) |>
  ggplot(aes(ratio, fct_reorder(language, ratio), fill = ratio > 1)) +
  geom_col() +
  geom_vline(xintercept = 1, color = "grey50", linetype = "dashed") +
  geom_text(aes(label = sprintf("×%.2f", ratio)), hjust = -0.1, size = 3) +
  scale_fill_manual(values = c("TRUE" = "#e34a33", "FALSE" = "#2c7fb8"),
                    labels = c("TRUE" = "Skews South",
                               "FALSE" = "Skews North"),
                    name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(
    title    = "How much more (or less) does the South use each language?",
    subtitle = "Ratio = (South share) ÷ (North share). >1 = Southern; <1 = Northern.",
    x = "Ratio", y = NULL
  ) +
  theme(legend.position = "bottom")
```

------------------------------------------------------------------------

## Plot 3 — Every country on the two axes

For each country: x-axis = % of its pushers on elite stacks (Rust + Go +
Swift + R), y-axis = % on maintenance stacks (PHP + Java + HTML + CSS).
Color = North / South. Labels on the most extreme countries on each axis.

The hypothesis predicts **Global-South countries cluster in the upper-left
(high maintenance, low elite)** and Global-North countries cluster in the
lower-right.

```{r fig-scatter, fig.width=10, fig.height=7}
label_set <- bind_rows(
  share |> slice_max(maint_share, n = 8),
  share |> slice_max(elite_share, n = 6),
  share |> slice_min(maint_share, n = 3)
) |> distinct(iso2_code, .keep_all = TRUE)

share_int <- share |>
  mutate(tooltip = paste0(ascii(country), " - ",
                          percent(maint_share, accuracy = 1)))

scatter_plot <- ggplot(share_int,
                       aes(elite_share, maint_share, color = hemisphere)) +
  geom_point_interactive(aes(size = total, tooltip = tooltip,
                             data_id = iso2_code),
                         alpha = 0.7) +
  geom_text_repel(data = label_set,
                  aes(label = country), size = 3, max.overlaps = 30,
                  segment.color = "grey60", segment.size = 0.3) +
  geom_hline(yintercept = mean(share$maint_share), color = "grey60",
             linetype = "dashed", linewidth = 0.3) +
  geom_vline(xintercept = mean(share$elite_share), color = "grey60",
             linetype = "dashed", linewidth = 0.3) +
  scale_x_continuous(labels = label_percent(accuracy = 1)) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  scale_size_area(max_size = 9, labels = label_comma(), name = "Pushers") +
  scale_color_manual(values = c("Global North" = "#2c7fb8",
                                "Global South" = "#e34a33"),
                     name = NULL) +
  labs(
    title    = "Every country on the two axes: elite vs. maintenance",
    subtitle = "Hypothesis predicts Global-South dots upper-left, Global-North dots lower-right. Hover for country + maintenance share.",
    x = "Share on elite stacks (Rust + Go + Swift + R)",
    y = "Share on maintenance stacks (PHP + Java + HTML + CSS)"
  ) +
  theme(legend.position = "bottom")

girafe(
  ggobj = scatter_plot,
  width_svg = 10, height_svg = 7,
  options = list(
    opts_tooltip(css = "background-color:#1a1a1a;color:white;padding:4px 8px;border-radius:4px;font-family:sans-serif;font-size:12px;"),
    opts_hover(css = "stroke:#111111;stroke-width:1.2px;")
  )
)
```

A summary number — the **separation** between the two clouds:

```{r}
share |>
  group_by(hemisphere) |>
  summarise(median_maint = percent(median(maint_share), 0.1),
            median_elite = percent(median(elite_share), 0.1),
            countries    = n(),
            .groups      = "drop")
```

------------------------------------------------------------------------

## Plot 4 — Per-topic "South lift"

Languages are blunt. Topics are sharp. For each suspect topic, compute the
**South-share** (% of South pushers on that topic, out of all South topic
activity) and the **North-share**, then plot `South / North`:

-   ratio > 1 → topic over-indexes in the South ("more Southern" than the
    average topic)
-   ratio < 1 → topic over-indexes in the North

Topics picked to span the maintenance / elite spectrum, with some
neutral controls.

```{r}
maint_topics <- c("wordpress", "woocommerce", "jquery", "laravel", "drupal",
                  "bootstrap", "magento", "php", "codeigniter", "joomla",
                  "android", "spring-boot", "html", "css", "wordpress-theme",
                  "wordpress-plugin")

elite_topics <- c("rust", "kubernetes", "llm", "machine-learning",
                  "deep-learning", "tensorflow", "pytorch", "golang",
                  "webassembly", "swift", "swiftui", "ios", "scala",
                  "haskell", "openai", "langchain")

control_topics <- c("python", "docker", "react", "javascript", "typescript")

all_topics <- unique(c(maint_topics, elite_topics, control_topics))
```

```{r fig-lift, fig.width=10, fig.height=10}
hem_totals <- topics_snap |>
  group_by(hemisphere) |>
  summarise(hem_total = sum(num_pushers), .groups = "drop")

lift <- topics_snap |>
  filter(topic %in% all_topics) |>
  group_by(hemisphere, topic) |>
  summarise(pushers = sum(num_pushers), .groups = "drop") |>
  inner_join(hem_totals, by = "hemisphere") |>
  mutate(share = pushers / hem_total) |>
  select(hemisphere, topic, share) |>
  pivot_wider(names_from = hemisphere, values_from = share,
              values_fill = NA_real_) |>
  filter(!is.na(`Global North`), !is.na(`Global South`)) |>
  mutate(ratio = `Global South` / `Global North`,
         bucket = case_when(
           topic %in% maint_topics ~ "Maintenance-suspect",
           topic %in% elite_topics ~ "Elite-suspect",
           TRUE                    ~ "Control"
         ),
         bucket = factor(bucket, levels = c("Maintenance-suspect",
                                            "Control", "Elite-suspect")))

ggplot(lift, aes(ratio, fct_reorder(topic, ratio), fill = bucket)) +
  geom_col() +
  geom_vline(xintercept = 1, color = "grey40", linetype = "dashed") +
  geom_text(aes(label = sprintf("×%.2f", ratio)),
            hjust = if_else(lift$ratio < 1, 1.1, -0.1),
            size = 3) +
  scale_x_log10(breaks = c(0.1, 0.25, 0.5, 1, 2, 4, 10),
                labels = c("×0.1", "×0.25", "×0.5", "×1",
                           "×2", "×4", "×10")) +
  scale_fill_manual(values = c("Maintenance-suspect" = "#e34a33",
                               "Elite-suspect"       = "#2c7fb8",
                               "Control"             = "grey60"),
                    name = NULL) +
  labs(
    title    = "Topic-level South lift: which work over-indexes where?",
    subtitle = "Bar = (South share of South pushers) ÷ (North share of North pushers). Right = Southern; Left = Northern.",
    x = "South / North ratio (log scale)", y = NULL
  ) +
  theme(legend.position = "bottom")
```

------------------------------------------------------------------------

## Verdict

```{r verdict, echo=FALSE}
ns_summary <- share |>
  group_by(hemisphere) |>
  summarise(median_maint = median(maint_share),
            median_elite = median(elite_share),
            .groups = "drop")
ns_summary
```

What the four reads above show:

1.  **The maps**: maintenance-stack share is visibly higher across most
    of Sub-Saharan Africa, South Asia, the Arab States, and Latin America
    than in North America, Western Europe, or the rich East Asian
    economies. The color difference between Plot-1 panels isn't subtle.
2.  **Head-to-head bars**: PHP, Java, HTML, CSS each have a Southern
    share above the Northern share. Rust, Go, Swift, R are the inverse —
    Northern share above Southern.
3.  **Country scatter**: Global-South dots cluster in the upper portion
    of the plot (more maintenance) and the left portion (less elite).
    The split is messy at the boundary but the cluster centers are
    clearly offset.
4.  **Topic lift**: WordPress / Laravel / jQuery / PHP / Drupal /
    Magento all sit firmly on the right (Southern); Rust / Kubernetes /
    LLM / SwiftUI / Haskell sit firmly on the left (Northern). The
    "controls" (python, docker, react) sit near ×1 as expected.

**Hypothesis: supported on the data.** Global-South GitHub activity
disproportionately concentrates on the high-demand maintenance stack
(CMS-era PHP, enterprise Java, frontend templating), and Global-North
activity disproportionately concentrates on newer, narrower-hiring
stacks (Rust, Go, Swift, ML / LLM tooling).

**Caveats worth flagging in a polished writeup**:

-   GitHub's 100-developer privacy floor systematically thins out
    low-volume topics in low-HDI economies (documented in
    `topics_by_hdi.qmd`), which can both *exaggerate* the maintenance
    skew (small "elite" cells get suppressed) and *attenuate* the
    headline difference (many South countries drop out entirely).
-   The bucket labels are editorial. A reader who thinks of Java
    primarily as Android development, or of Swift as Apple-fan
    parochialism, will read the plot differently. The buckets are
    explicit so readers can second-guess them.
-   Open-source push counts are not labor markets. Outsourced PHP
    *work* in Manila may never appear on public GitHub. What this proves
    is that the *open-source pattern* is consistent with the
    outsourcing hypothesis — not that the outsourcing itself is
    visible.
