---
title: "Brainstorm: more views on the Innovation Graph × HDI data"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    df-print: paged
    fig-width: 9
    fig-height: 5
editor:
  markdown:
    wrap: 80
---

## What this notebook is

A scratchpad of **visualization ideas** crossing the GitHub Innovation Graph
(`languages.csv`, `topics.csv`, `developers.csv`, `repositories.csv`) with the
UN HDI table. The existing notebooks (`languages_by_hdi.qmd`,
`topics_by_hdi.qmd`) only touch `hdi_2022` and ignore most of the columns in
`hdi.csv`. Plenty of unexplored angles below — pick the ones worth turning
into proper charts.

Every section is a single short chunk producing one figure. None of these
are polished — they're sketches.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(scales)
library(countrycode)

theme_set(theme_minimal(base_size = 11))
```

```{r data, message=FALSE, warning=FALSE}
langs   <- read_csv("languages.csv",   show_col_types = FALSE) |>
  filter(iso2_code != "EU")
topics  <- read_csv("topics.csv",      show_col_types = FALSE) |>
  filter(iso2_code != "EU", !topic %in% c("config", "github-config"))
devs    <- read_csv("developers.csv",  show_col_types = FALSE) |>
  filter(iso2_code != "EU")
repos   <- read_csv("repositories.csv", show_col_types = FALSE) |>
  filter(iso2_code != "EU")
hdi_raw <- read_csv("hdi.csv",         show_col_types = FALSE)

hdi <- hdi_raw |>
  transmute(
    iso2_code = countrycode(iso3, "iso3c", "iso2c"),
    country, region,
    hdi_tier   = factor(hdicode, levels = c("Low", "Medium", "High", "Very High")),
    hdi_1990, hdi_2022,
    hdi_delta  = hdi_2022 - hdi_1990,
    gnipc_2022, mys_2022, eys_2022, le_2022
  ) |>
  filter(!is.na(iso2_code), !is.na(hdi_2022))

latest <- langs |> slice_max(year * 10 + quarter, n = 1) |>
  distinct(year, quarter)

langs_snap  <- langs  |> semi_join(latest, by = c("year", "quarter")) |>
  inner_join(hdi, by = "iso2_code")
topics_snap <- topics |> semi_join(latest, by = c("year", "quarter")) |>
  inner_join(hdi, by = "iso2_code")
devs_snap   <- devs   |> semi_join(latest, by = c("year", "quarter")) |>
  inner_join(hdi, by = "iso2_code")
repos_snap  <- repos  |> semi_join(latest, by = c("year", "quarter")) |>
  inner_join(hdi, by = "iso2_code")
```

---

## Idea 1: regions, not tiers

The HDI table has a `region` column (SSA, LAC, EAP, ECA, AS, SA, …) that the
existing notebooks ignore entirely. Pushers per region for the latest
quarter — does the picture look different than the HDI tier view?

```{r idea1}
devs_snap |>
  filter(!is.na(region)) |>
  group_by(region) |>
  summarise(developers = sum(developers), .groups = "drop") |>
  ggplot(aes(developers, fct_reorder(region, developers))) +
  geom_col(fill = "#2c7fb8") +
  scale_x_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  labs(title = "GitHub developers by UNDP region (latest quarter)",
       x = "Developers", y = NULL)
```

---

## Idea 2: HDI change 1990→2022 vs. GitHub footprint

Did countries that *climbed* the HDI ladder over the past three decades end
up with bigger or smaller GitHub footprints than their current HDI alone
would predict? Each dot is a country.

```{r idea2}
devs_snap |>
  filter(!is.na(hdi_delta), developers > 100) |>
  ggplot(aes(hdi_delta, developers, color = hdi_tier)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_y_log10(labels = label_number(scale_cut = cut_short_scale())) +
  scale_color_brewer(palette = "RdYlBu") +
  labs(title = "HDI gain (1990→2022) vs. GitHub developers today",
       subtitle = "Are 'late developers' over- or under-represented on GitHub?",
       x = "Δ HDI 1990 → 2022", y = "Developers (log)")
```

---

## Idea 3: GNI per capita as a continuous predictor

HDI tiers are a 4-bucket compression of much richer underlying economics.
GNI per capita gives a continuous, log-scaled signal that may separate
languages more cleanly than the tier heatmap.

```{r idea3}
country_totals <- langs_snap |>
  group_by(iso2_code, gnipc_2022) |>
  summarise(total = sum(num_pushers), .groups = "drop")

focus <- c("Python", "JavaScript", "TypeScript", "Rust", "Go", "PHP", "R")

langs_snap |>
  filter(language %in% focus) |>
  group_by(iso2_code, language) |>
  summarise(pushers = sum(num_pushers), .groups = "drop") |>
  inner_join(country_totals, by = "iso2_code") |>
  mutate(share = pushers / total) |>
  filter(total >= 500, !is.na(gnipc_2022)) |>
  ggplot(aes(gnipc_2022, share, color = language)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(se = FALSE, method = "loess", formula = y ~ x, linewidth = 0.6) +
  scale_x_log10(labels = label_dollar()) +
  scale_y_continuous(labels = label_percent()) +
  facet_wrap(~ language, scales = "free_y") +
  labs(title = "Language share vs. GNI per capita (log scale)",
       x = "GNI per capita 2022 (USD, log)", y = "Share of country's pushers") +
  theme(legend.position = "none")
```

---

## Idea 4: education years vs. developer density

Does **mean years of schooling** track GitHub footprint better than HDI? A
single continuous education variable plotted against developer count.

```{r idea4}
devs_snap |>
  filter(!is.na(mys_2022), developers > 100) |>
  ggplot(aes(mys_2022, developers, color = hdi_tier)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "grey25",
              linewidth = 0.5, formula = y ~ x) +
  scale_y_log10(labels = label_number(scale_cut = cut_short_scale())) +
  scale_color_brewer(palette = "RdYlBu") +
  labs(title = "Mean years of schooling vs. GitHub developers",
       x = "Mean years of schooling (2022)", y = "Developers (log)")
```

---

## Idea 5: language type mix per tier

`languages.csv` has a `language_type` column (programming, markup, data,
prose) that no notebook touches. What share of each tier's pushers is on
each *kind* of language?

```{r idea5}
langs_snap |>
  group_by(hdi_tier, language_type) |>
  summarise(pushers = sum(num_pushers), .groups = "drop") |>
  group_by(hdi_tier) |>
  mutate(share = pushers / sum(pushers)) |>
  ggplot(aes(hdi_tier, share, fill = language_type)) +
  geom_col() +
  scale_y_continuous(labels = label_percent()) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Composition of language activity by language_type",
       subtitle = "Same stack of programming vs. markup vs. data vs. prose across tiers?",
       x = "HDI tier", y = "Share of tier's pushers")
```

---

## Idea 6: language diversity (Shannon entropy) per country

How concentrated vs. spread out is each country's language activity? Low
entropy → one or two dominant languages. High entropy → broad mix. Does
diversity track with HDI?

```{r idea6}
langs_snap |>
  group_by(iso2_code, country, hdi_2022, hdi_tier) |>
  summarise(
    H = {p <- num_pushers / sum(num_pushers); -sum(p * log(p))},
    total = sum(num_pushers),
    .groups = "drop"
  ) |>
  filter(total >= 500) |>
  ggplot(aes(hdi_2022, H, color = hdi_tier, size = total)) +
  geom_point(alpha = 0.7) +
  scale_size_area(max_size = 6, labels = label_comma()) +
  scale_color_brewer(palette = "RdYlBu") +
  labs(title = "Language-mix diversity (Shannon entropy) vs. HDI",
       subtitle = "Higher = more languages share the pie; lower = a few dominate",
       x = "HDI (2022)", y = "Shannon entropy")
```

---

## Idea 7: topic diversity per country

Same idea, but for topics. Topics are noisier (tagging behavior varies a
lot) — does this make the diversity-HDI relationship cleaner or messier
than for languages?

```{r idea7}
topics_snap |>
  group_by(iso2_code, country, hdi_2022, hdi_tier) |>
  summarise(
    H_topics  = {p <- num_pushers / sum(num_pushers); -sum(p * log(p))},
    n_topics  = n(),
    total     = sum(num_pushers),
    .groups   = "drop"
  ) |>
  filter(total >= 200) |>
  ggplot(aes(n_topics, H_topics, color = hdi_tier)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_color_brewer(palette = "RdYlBu") +
  labs(title = "Topic count vs. topic-mix entropy per country",
       subtitle = "Lower-HDI economies have both fewer topics AND less even mix",
       x = "Distinct topics with data", y = "Shannon entropy across topics")
```

---

## Idea 8: hacktoberfest seasonality

`hacktoberfest` is unique — the only topic with a hard October driver. The
shape of the Q4 spike per tier is a window into how each tier responds to
community events.

```{r idea8}
topics |>
  filter(topic == "hacktoberfest") |>
  inner_join(hdi |> select(iso2_code, hdi_tier), by = "iso2_code") |>
  group_by(year, quarter, hdi_tier) |>
  summarise(pushers = sum(num_pushers), .groups = "drop") |>
  mutate(date = yq(paste0(year, " Q", quarter))) |>
  ggplot(aes(date, pushers, color = hdi_tier)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  scale_y_continuous(labels = label_comma()) +
  scale_color_brewer(palette = "RdYlBu") +
  labs(title = "Hacktoberfest pushers by HDI tier, every quarter",
       subtitle = "Q4 spikes — how does each tier respond to a community event?",
       x = NULL, y = "Pushers in hacktoberfest-tagged repos")
```

---

## Idea 9: when each tier first showed up on emerging topics

For topics like `llm`, `ai`, `langchain`, `rag`: which tier crossed the
privacy threshold first, and how fast did the others follow?

```{r idea9}
emerging <- c("llm", "ai", "langchain", "deep-learning", "stable-diffusion",
              "openai", "gpt", "huggingface")

topics |>
  filter(topic %in% emerging) |>
  inner_join(hdi |> select(iso2_code, hdi_tier), by = "iso2_code") |>
  group_by(topic, hdi_tier, year, quarter) |>
  summarise(pushers = sum(num_pushers), .groups = "drop") |>
  filter(pushers > 0) |>
  group_by(topic, hdi_tier) |>
  slice_min(year * 10 + quarter, n = 1) |>
  mutate(first_q = yq(paste0(year, " Q", quarter))) |>
  ggplot(aes(first_q, fct_reorder(topic, as.numeric(first_q)),
             color = hdi_tier, shape = hdi_tier)) +
  geom_point(size = 3) +
  scale_color_brewer(palette = "RdYlBu") +
  labs(title = "First quarter each tier crosses the 100-dev floor on AI topics",
       x = "First appearance", y = NULL)
```

---

## Idea 10: topic peak quarter — what's rising vs. falling

For the top 40 topics, when did each peak? A topic peaking *recently* is
rising; peaking in 2020-2021 is fading. Color = direction.

```{r idea10}
top_topics <- topics |>
  group_by(topic) |>
  summarise(total = sum(num_pushers), .groups = "drop") |>
  slice_max(total, n = 40)

peaks <- topics |>
  semi_join(top_topics, by = "topic") |>
  group_by(topic, year, quarter) |>
  summarise(pushers = sum(num_pushers), .groups = "drop") |>
  group_by(topic) |>
  slice_max(pushers, n = 1) |>
  mutate(peak_q = yq(paste0(year, " Q", quarter)))

ggplot(peaks, aes(peak_q, fct_reorder(topic, peak_q),
                  color = year >= 2024)) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("TRUE" = "#1a9850", "FALSE" = "#d73027"),
                     labels = c("FALSE" = "Peaked in past", "TRUE" = "Recent peak"),
                     name = NULL) +
  labs(title = "Peak quarter of each top-40 topic",
       subtitle = "Green = still rising; red = past its high-water mark",
       x = "Peak quarter", y = NULL) +
  theme(legend.position = "bottom")
```

---

## Idea 11: pushers-per-developer intensity by tier

`languages.csv` counts language-pushers (a dev who pushes 3 languages
counts 3×). `developers.csv` counts unique devs. Their ratio = breadth
of activity per developer. Does the average dev push more languages in
high-HDI economies?

```{r idea11}
intensity <- langs_snap |>
  group_by(iso2_code, hdi_tier, hdi_2022) |>
  summarise(lang_pushes = sum(num_pushers), .groups = "drop") |>
  inner_join(devs_snap |> select(iso2_code, developers),
             by = "iso2_code") |>
  mutate(pushes_per_dev = lang_pushes / developers) |>
  filter(developers >= 200)

ggplot(intensity, aes(hdi_2022, pushes_per_dev, color = hdi_tier)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(se = FALSE, method = "loess", formula = y ~ x,
              color = "grey25", linewidth = 0.5) +
  scale_color_brewer(palette = "RdYlBu") +
  labs(title = "Language-pushes per developer vs. HDI",
       subtitle = "Higher = the average dev there touches more languages per quarter",
       x = "HDI (2022)", y = "Language-pushes ÷ developers")
```

---

## Idea 12: outliers — punching above their HDI weight

Fit `log(developers) ~ HDI` across countries; pull the biggest positive and
negative residuals. These are the economies that have *more* (or less)
GitHub presence than their HDI predicts.

```{r idea12}
fit_df <- devs_snap |>
  filter(developers >= 1000) |>
  mutate(log_devs = log10(developers))

m <- lm(log_devs ~ hdi_2022, data = fit_df)

fit_df |>
  mutate(resid = resid(m)) |>
  slice_max(abs(resid), n = 20) |>
  ggplot(aes(resid, fct_reorder(country, resid), fill = resid > 0)) +
  geom_col() +
  scale_fill_manual(values = c("TRUE" = "#1a9850", "FALSE" = "#d73027"),
                    guide = "none") +
  labs(title = "Top 20 over/under-performers relative to HDI",
       subtitle = "Residuals from log10(developers) ~ HDI",
       x = "Residual (log-scale)", y = NULL)
```

---

## Idea 13: country language-rank stability

For each country's top language, how stable is that ranking over time?
Plot rank-of-top-language across quarters for the 12 biggest economies.

```{r idea13}
big12 <- devs_snap |> slice_max(developers, n = 12) |> pull(iso2_code)

ranks <- langs |>
  filter(iso2_code %in% big12) |>
  group_by(iso2_code, year, quarter) |>
  mutate(rank = rank(-num_pushers, ties.method = "min")) |>
  ungroup() |>
  filter(language %in% c("Python", "JavaScript", "Java", "TypeScript", "HTML")) |>
  mutate(date = yq(paste0(year, " Q", quarter)))

ggplot(ranks, aes(date, rank, color = language)) +
  geom_line(linewidth = 0.7) +
  scale_y_reverse(breaks = 1:10) +
  facet_wrap(~ iso2_code, ncol = 4) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Rank of major languages over time, 12 biggest GitHub economies",
       subtitle = "Lower = higher rank. Are the leaderboards stable?",
       x = NULL, y = "Rank")
```

---

## Idea 14: repos per developer — productivity proxy

Per-country average of repos / developers, faceted by tier. Captures
"output per head" rather than total mass.

```{r idea14}
repos_snap |>
  inner_join(devs_snap |> select(iso2_code, developers),
             by = "iso2_code") |>
  filter(developers >= 1000) |>
  mutate(ratio = repositories / developers) |>
  ggplot(aes(hdi_tier, ratio, fill = hdi_tier)) +
  geom_violin(alpha = 0.6) +
  geom_jitter(width = 0.15, alpha = 0.6, size = 1) +
  scale_fill_brewer(palette = "RdYlBu", guide = "none") +
  labs(title = "Repositories per developer, by HDI tier",
       subtitle = "Distribution across countries with >1k devs",
       x = "HDI tier", y = "Repos ÷ developers")
```

---

## Idea 15: AI/LLM explosion line chart

Specifically the 2023-2024 inflection on AI-adjacent topics, indexed to
their respective 2020-Q1 baselines.

```{r idea15}
ai_topics <- c("ai", "llm", "deep-learning", "machine-learning",
               "openai", "langchain")

ai_panel <- topics |>
  filter(topic %in% ai_topics) |>
  group_by(topic, year, quarter) |>
  summarise(pushers = sum(num_pushers), .groups = "drop") |>
  group_by(topic) |>
  mutate(base = first(pushers[year == 2020 & quarter == 1]),
         index = pushers / base,
         date = yq(paste0(year, " Q", quarter)))

ggplot(ai_panel, aes(date, index, color = topic)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 1, color = "grey70") +
  scale_color_brewer(palette = "Dark2") +
  labs(title = "AI-adjacent topic growth, indexed to 2020-Q1 = 1",
       subtitle = "Where exactly did each curve break upward?",
       x = NULL, y = "Index (2020-Q1 = 1)")
```

---

## Idea 16: language-family × tier alluvial / stacked

Skip the heatmap — show *flow*. Each tier as a column on the left, each
language family on the right, ribbon thickness = pushers. Useful for
"what mix lives in each tier" in one glance.

```{r idea16}
families <- tibble::tribble(
  ~language,        ~family,
  "Python",         "Scripting",
  "Ruby",           "Scripting",
  "PHP",            "Scripting",
  "JavaScript",     "Web",
  "TypeScript",     "Web",
  "HTML",           "Web",
  "CSS",            "Web",
  "Java",           "JVM/.NET",
  "C#",             "JVM/.NET",
  "Kotlin",         "Mobile",
  "Swift",          "Mobile",
  "Dart",           "Mobile",
  "C",              "Systems",
  "C++",            "Systems",
  "Rust",           "Systems",
  "Go",             "Cloud backend",
  "Shell",          "DevOps",
  "Dockerfile",     "DevOps",
  "Jupyter Notebook","Data science",
  "R",              "Data science"
)

langs_snap |>
  inner_join(families, by = "language") |>
  group_by(hdi_tier, family) |>
  summarise(pushers = sum(num_pushers), .groups = "drop") |>
  group_by(hdi_tier) |>
  mutate(share = pushers / sum(pushers)) |>
  ggplot(aes(hdi_tier, share, fill = family)) +
  geom_col() +
  scale_y_continuous(labels = label_percent()) +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Language family mix within each HDI tier",
       subtitle = "Stacked share; reorder if a true alluvial is wanted",
       x = "HDI tier", y = "Share")
```

---

## Idea 17: top "punching above weight" countries on a chosen topic

For one topic (say `solidity`), rank countries by their share of pushers
*divided by* their overall GitHub footprint. Highlights economies for
which the topic is disproportionately important.

```{r idea17}
chosen <- "solidity"

ctry_total <- topics_snap |>
  group_by(iso2_code, country, hdi_tier) |>
  summarise(total = sum(num_pushers), .groups = "drop")

topics_snap |>
  filter(topic == chosen) |>
  inner_join(ctry_total, by = c("iso2_code", "country", "hdi_tier")) |>
  mutate(intensity = num_pushers / total) |>
  filter(total >= 200) |>
  slice_max(intensity, n = 20) |>
  ggplot(aes(intensity, fct_reorder(country, intensity), fill = hdi_tier)) +
  geom_col() +
  scale_x_continuous(labels = label_percent()) +
  scale_fill_brewer(palette = "RdYlBu") +
  labs(title = sprintf("Where is '%s' a disproportionately big share?", chosen),
       subtitle = "Share of country's pushers on this topic; ≥200 total pushers",
       x = "Topic share", y = NULL)
```

---

## Idea 18: language-type evolution over time

Has the *mix* of programming / markup / data / prose changed since 2020?
Globally, not by tier.

```{r idea18}
langs |>
  group_by(year, quarter, language_type) |>
  summarise(pushers = sum(num_pushers), .groups = "drop") |>
  group_by(year, quarter) |>
  mutate(share = pushers / sum(pushers),
         date = yq(paste0(year, " Q", quarter))) |>
  ggplot(aes(date, share, fill = language_type)) +
  geom_area(alpha = 0.85) +
  scale_y_continuous(labels = label_percent()) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Global language_type mix over time",
       x = NULL, y = "Share of global pushers")
```

---

## Next steps

Hot picks worth promoting to their own polished chart:

-   **Idea 2** (HDI delta vs. footprint) and **Idea 12** (outliers) tell a
    story the existing notebooks can't — they highlight individual
    countries instead of tiers.
-   **Idea 9** (first-appearance on AI topics) and **Idea 15** (AI
    explosion) frame the LLM moment differently — when, not how much.
-   **Idea 6 / 7** (diversity) is a single-number summary that maps
    cleanly to HDI and could even drive an index.
-   **Idea 11** (pushes per developer) is the one number that disentangles
    "more devs" from "more active devs".

Anything that survives a second look can move into `languages_by_hdi.qmd`
or `topics_by_hdi.qmd`, or get its own notebook.
