---
title: "GitHub Languages by Economic Development (HDI)"
format:
  html:
    embed-resources: true
    toc: true
    df-print: paged
    fig-width: 9
    fig-height: 6
editor:
  markdown:
    wrap: 80
---

## Idea

Combine the [GitHub Innovation Graph](https://github.com/github/innovationgraph)
language panel with the **UN Human Development Index** to ask: *do high-HDI
and low-HDI economies use different languages, and are the trends moving in
different directions?*

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(scales)
library(countrycode)

theme_set(theme_minimal(base_size = 11))
```

## Data

```{r}
langs <- read_csv("languages.csv", show_col_types = FALSE)

hdi_raw <- read_csv("hdi.csv", show_col_types = FALSE)
```

The HDI file uses ISO3 codes; the GitHub file uses ISO2. We'll add an ISO2
column to the HDI data and keep the most recent year (2022 in this release).

```{r}
hdi <- hdi_raw |>
  transmute(
    iso2_code = countrycode(iso3, "iso3c", "iso2c"),
    country   = country,
    region    = region,
    hdi_tier  = factor(hdicode, levels = c("Low", "Medium", "High", "Very High")),
    hdi       = hdi_2022
  ) |>
  filter(!is.na(iso2_code), !is.na(hdi))

glimpse(hdi)
```

How many GitHub-data countries match an HDI value?

```{r}
gh_countries <- langs |> distinct(iso2_code)
inner_join(gh_countries, hdi, by = "iso2_code") |> nrow()
```

The handful of unmatched economies are mostly small territories (e.g., Hong
Kong is reported separately by GitHub but folded into China by UNDP). We
proceed with the matched set.

## Snapshot quarter

```{r}
latest <- langs |> slice_max(year * 10 + quarter, n = 1) |>
  distinct(year, quarter)
latest

snap <- langs |>
  semi_join(latest, by = c("year", "quarter")) |>
  inner_join(hdi, by = "iso2_code")
```

## Language share within each HDI tier

For each tier we sum pushers across all member countries, then take each
language's share of that tier's total.

```{r}
focus_langs <- c("Python", "JavaScript", "TypeScript", "Java", "C++",
                 "C#", "Go", "Rust", "PHP", "Ruby",
                 "HTML", "Shell", "Kotlin", "Swift", "R")

tier_totals <- snap |>
  group_by(hdi_tier) |>
  summarise(tier_total = sum(num_pushers), .groups = "drop")

tier_lang <- snap |>
  filter(language %in% focus_langs) |>
  group_by(hdi_tier, language) |>
  summarise(pushers = sum(num_pushers), .groups = "drop") |>
  inner_join(tier_totals, by = "hdi_tier") |>
  mutate(share = pushers / tier_total) |>
  mutate(text_color = if_else(share > 0.5 * max(share), "white", "grey15"))
```

```{r fig-tier-heatmap, fig.width=9, fig.height=5}
ggplot(tier_lang,
       aes(hdi_tier, fct_reorder(language, share, .fun = max), fill = share)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = percent(share, accuracy = 0.1), color = text_color),
            size = 2.8) +
  scale_color_identity() +
  scale_fill_viridis_c(option = "mako", direction = -1, end = 0.85,
                       labels = label_percent(),
                       name = "Share of\ntier's pushers") +
  labs(
    title    = "Language share within each HDI tier",
    subtitle = "Each cell = % of pushers in that tier using that language (latest quarter)",
    x = "HDI tier", y = NULL
  )
```

## HDI vs language share, country by country

Here every dot is a country: x = HDI, y = share of that country's pushers
using the language. The smoothed line is a LOESS fit weighted by total
pushers, so big economies pull more.

```{r fig-scatter, fig.width=10, fig.height=8}
country_totals <- snap |>
  group_by(iso2_code, hdi, hdi_tier) |>
  summarise(total = sum(num_pushers), .groups = "drop")

scat <- snap |>
  filter(language %in% focus_langs) |>
  group_by(iso2_code, language) |>
  summarise(pushers = sum(num_pushers), .groups = "drop") |>
  inner_join(country_totals, by = "iso2_code") |>
  mutate(share = pushers / total) |>
  filter(total >= 500)  # drop tiny economies that produce noisy shares

ggplot(scat, aes(hdi, share)) +
  geom_point(aes(size = total, color = hdi_tier), alpha = 0.6) +
  geom_smooth(aes(weight = total), method = "loess", se = FALSE,
              color = "grey25", linewidth = 0.6, formula = y ~ x) +
  facet_wrap(~ language, scales = "free_y") +
  scale_size_area(max_size = 6, labels = label_comma(),
                  name = "Pushers") +
  scale_y_continuous(labels = label_percent()) +
  scale_color_brewer(palette = "RdYlBu", name = "HDI tier") +
  labs(
    title    = "Language share vs. HDI",
    subtitle = "One point per country (≥500 pushers); line = pusher-weighted LOESS",
    x = "HDI (2022)", y = "Share of country's pushers"
  ) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))
```

## Where each language sits on the development gradient

To compress the picture into one chart: for each focus language compute the
*pusher-weighted mean HDI* of its users.

```{r fig-mean-hdi, fig.width=8, fig.height=5}
lang_meanhdi <- snap |>
  filter(language %in% focus_langs) |>
  group_by(language) |>
  summarise(
    mean_hdi = weighted.mean(hdi, num_pushers),
    pushers  = sum(num_pushers),
    .groups  = "drop"
  )

global_mean <- weighted.mean(snap$hdi, snap$num_pushers)

ggplot(lang_meanhdi,
       aes(mean_hdi, fct_reorder(language, mean_hdi),
           color = mean_hdi)) +
  geom_vline(xintercept = global_mean, linetype = "dashed",
             color = "grey50") +
  geom_segment(aes(x = global_mean, xend = mean_hdi, yend = language),
               linewidth = 0.6) +
  geom_point(size = 3) +
  scale_color_viridis_c(option = "mako", direction = -1, guide = "none") +
  annotate("text", x = global_mean, y = 0.6,
           label = sprintf("Global mean (%.3f)", global_mean),
           hjust = -0.05, vjust = 0, size = 3, color = "grey40") +
  labs(
    title    = "Pusher-weighted average HDI of each language's users",
    subtitle = "Right of dashed line = skews toward higher-HDI economies",
    x = "Mean HDI", y = NULL
  )
```

## Trends within tiers, 2020 → today

Are some tiers growing faster on certain languages? We index each tier's
quarterly pusher counts to its 2020-Q1 baseline.

```{r fig-trend-tier, fig.width=10, fig.height=7}
panel <- langs |>
  inner_join(hdi |> select(iso2_code, hdi_tier), by = "iso2_code") |>
  filter(language %in% focus_langs) |>
  group_by(year, quarter, hdi_tier, language) |>
  summarise(pushers = sum(num_pushers), .groups = "drop") |>
  mutate(date = yq(paste0(year, " Q", quarter)))

baseline <- panel |>
  filter(year == 2020, quarter == 1) |>
  select(hdi_tier, language, base = pushers)

panel <- panel |>
  inner_join(baseline, by = c("hdi_tier", "language")) |>
  mutate(index = pushers / base)

ggplot(panel, aes(date, index, color = hdi_tier)) +
  geom_hline(yintercept = 1, color = "grey70", linewidth = 0.3) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ language, scales = "free_y") +
  scale_color_brewer(palette = "RdYlBu", name = "HDI tier") +
  labs(
    title    = "Pusher growth by HDI tier, indexed to 2020-Q1 = 1",
    subtitle = "Higher line = faster growth in that tier",
    x = NULL, y = "Index (2020-Q1 = 1)"
  ) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))
```

## Total developer growth by tier

Same indexing, but summing *all* languages — i.e., raw developer-population
growth.

```{r fig-total-growth, fig.width=9, fig.height=5}
total_panel <- langs |>
  inner_join(hdi |> select(iso2_code, hdi_tier), by = "iso2_code") |>
  group_by(year, quarter, hdi_tier) |>
  summarise(pushers = sum(num_pushers), .groups = "drop") |>
  mutate(date = yq(paste0(year, " Q", quarter)))

total_base <- total_panel |>
  filter(year == 2020, quarter == 1) |>
  select(hdi_tier, base = pushers)

total_panel |>
  inner_join(total_base, by = "hdi_tier") |>
  mutate(index = pushers / base) |>
  ggplot(aes(date, index, color = hdi_tier)) +
  geom_hline(yintercept = 1, color = "grey70", linewidth = 0.3) +
  geom_line(linewidth = 1) +
  scale_color_brewer(palette = "RdYlBu", name = "HDI tier") +
  labs(
    title    = "GitHub pusher growth by HDI tier, indexed to 2020-Q1",
    x = NULL, y = "Index (2020-Q1 = 1)"
  )
```

## Pseudo-grouping: languages by what they're for

Rolling individual languages up into rough functional families lets us ask
"is this **kind** of work skewed by HDI?" instead of one language at a time.
Many languages span uses (Python is data *and* scripting; Java is JVM *and*
mobile-Android), so this is a primary-association call, not a clean
partition.

```{r}
lang_groups <- tribble(
  ~group,                ~language,
  # ---- Web (frontend) ----
  "Web frontend",        "JavaScript",
  "Web frontend",        "TypeScript",
  "Web frontend",        "HTML",
  "Web frontend",        "CSS",
  "Web frontend",        "SCSS",
  "Web frontend",        "Less",
  "Web frontend",        "Vue",
  "Web frontend",        "Svelte",
  "Web frontend",        "EJS",
  "Web frontend",        "Handlebars",
  "Web frontend",        "Blade",
  "Web frontend",        "Smarty",
  "Web frontend",        "Pug",

  # ---- Mobile ----
  "Mobile",              "Swift",
  "Mobile",              "Objective-C",
  "Mobile",              "Objective-C++",
  "Mobile",              "Kotlin",
  "Mobile",              "Dart",

  # ---- Systems / low-level ----
  "Systems",             "C",
  "Systems",             "C++",
  "Systems",             "Rust",
  "Systems",             "Assembly",
  "Systems",             "Zig",

  # ---- Enterprise / JVM / .NET ----
  "Enterprise / JVM / .NET", "Java",
  "Enterprise / JVM / .NET", "C#",
  "Enterprise / JVM / .NET", "Scala",
  "Enterprise / JVM / .NET", "F#",
  "Enterprise / JVM / .NET", "Groovy",
  "Enterprise / JVM / .NET", "Visual Basic .NET",

  # ---- Cloud-native backend ----
  "Cloud backend",       "Go",
  "Cloud backend",       "Elixir",
  "Cloud backend",       "Erlang",

  # ---- Scripting / general-purpose ----
  "Scripting",           "Python",
  "Scripting",           "Ruby",
  "Scripting",           "PHP",
  "Scripting",           "Perl",
  "Scripting",           "Lua",
  "Scripting",           "Hack",

  # ---- Data science / scientific ----
  "Data science",        "Jupyter Notebook",
  "Data science",        "R",
  "Data science",        "MATLAB",
  "Data science",        "Julia",
  "Data science",        "Stata",
  "Data science",        "SAS",

  # ---- DevOps / shell / config ----
  "DevOps / config",     "Shell",
  "DevOps / config",     "PowerShell",
  "DevOps / config",     "Dockerfile",
  "DevOps / config",     "Makefile",
  "DevOps / config",     "CMake",
  "DevOps / config",     "HCL",
  "DevOps / config",     "Batchfile",
  "DevOps / config",     "Nix",
  "DevOps / config",     "Starlark",
  "DevOps / config",     "Procfile",

  # ---- Game / graphics ----
  "Game / graphics",     "GLSL",
  "Game / graphics",     "HLSL",
  "Game / graphics",     "ShaderLab",
  "Game / graphics",     "GDScript",

  # ---- Smart contracts ----
  "Smart contracts",     "Solidity",
  "Smart contracts",     "Vyper"
) |>
  mutate(group = factor(group, levels = c(
    "Web frontend", "Mobile", "Systems", "Enterprise / JVM / .NET",
    "Cloud backend", "Scripting", "Data science",
    "DevOps / config", "Game / graphics", "Smart contracts"
  )))
```

Sanity check — what fraction of all pushers do these groups cover?

```{r}
grouped <- snap |> inner_join(lang_groups, by = "language")

snap |>
  summarise(grouped   = sum(num_pushers[language %in% lang_groups$language]),
            total     = sum(num_pushers)) |>
  mutate(coverage = percent(grouped / total, accuracy = 0.1))
```

### Group share within each HDI tier

```{r fig-grp-heatmap, fig.width=8, fig.height=5}
tier_totals <- snap |>
  group_by(hdi_tier) |>
  summarise(tier_total = sum(num_pushers), .groups = "drop")

tier_grp <- grouped |>
  group_by(hdi_tier, group) |>
  summarise(pushers = sum(num_pushers), .groups = "drop") |>
  inner_join(tier_totals, by = "hdi_tier") |>
  mutate(share = pushers / tier_total) |>
  mutate(text_color = if_else(share > 0.5 * max(share), "white", "grey15"))

ggplot(tier_grp,
       aes(hdi_tier, fct_rev(group), fill = share)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = percent(share, accuracy = 0.1), color = text_color),
            size = 3) +
  scale_color_identity() +
  scale_fill_viridis_c(option = "mako", direction = -1, end = 0.85,
                       labels = label_percent(),
                       name = "Share of\ntier's pushers") +
  labs(
    title    = "Share of each HDI tier's pushers, by language family",
    subtitle = "Columns sum to 100% across all groups (incl. uncategorized)",
    x = "HDI tier", y = NULL
  )
```

### Group share inside the country: HDI vs. share

Every dot is a country; line is pusher-weighted LOESS. Slope = how much
this *kind* of work tracks development.

```{r fig-grp-scatter, fig.width=11, fig.height=8}
country_totals <- snap |>
  group_by(iso2_code, hdi, hdi_tier) |>
  summarise(total = sum(num_pushers), .groups = "drop")

grp_scat <- grouped |>
  group_by(iso2_code, group) |>
  summarise(pushers = sum(num_pushers), .groups = "drop") |>
  inner_join(country_totals, by = "iso2_code") |>
  mutate(share = pushers / total) |>
  filter(total >= 500)

ggplot(grp_scat, aes(hdi, share)) +
  geom_point(aes(size = total, color = hdi_tier), alpha = 0.55) +
  geom_smooth(aes(weight = total), method = "loess", se = FALSE,
              color = "grey25", linewidth = 0.6, formula = y ~ x) +
  facet_wrap(~ group, scales = "free_y", ncol = 4) +
  scale_size_area(max_size = 5, labels = label_comma(),
                  name = "Pushers") +
  scale_y_continuous(labels = label_percent()) +
  scale_color_brewer(palette = "RdYlBu", name = "HDI tier") +
  labs(
    title    = "Group share vs. HDI",
    subtitle = "Slope of the line = how much this kind of work skews with development",
    x = "HDI (2022)", y = "Share of country's pushers"
  ) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))
```

### Where each group sits on the development gradient

```{r fig-grp-meanhdi, fig.width=8, fig.height=4}
grp_meanhdi <- grouped |>
  group_by(group) |>
  summarise(
    mean_hdi = weighted.mean(hdi, num_pushers),
    pushers  = sum(num_pushers),
    .groups  = "drop"
  )

global_mean <- weighted.mean(snap$hdi, snap$num_pushers)

ggplot(grp_meanhdi,
       aes(mean_hdi, fct_reorder(group, mean_hdi),
           color = mean_hdi)) +
  geom_vline(xintercept = global_mean, linetype = "dashed",
             color = "grey50") +
  geom_segment(aes(x = global_mean, xend = mean_hdi, yend = group),
               linewidth = 0.6) +
  geom_point(size = 3.5) +
  scale_color_viridis_c(option = "mako", direction = -1, guide = "none") +
  annotate("text", x = global_mean, y = 0.6,
           label = sprintf("Global mean (%.3f)", global_mean),
           hjust = -0.05, vjust = 0, size = 3, color = "grey40") +
  labs(
    title    = "Pusher-weighted mean HDI of each language family",
    subtitle = "Right of dashed line = skews high-HDI; left = skews low-HDI",
    x = "Mean HDI", y = NULL
  )
```

### Indexed growth by group × HDI tier

```{r fig-grp-trend, fig.width=11, fig.height=8}
grp_panel <- langs |>
  inner_join(hdi |> select(iso2_code, hdi_tier), by = "iso2_code") |>
  inner_join(lang_groups, by = "language") |>
  group_by(year, quarter, hdi_tier, group) |>
  summarise(pushers = sum(num_pushers), .groups = "drop") |>
  mutate(date = yq(paste0(year, " Q", quarter)))

grp_base <- grp_panel |>
  filter(year == 2020, quarter == 1, pushers > 0) |>
  select(hdi_tier, group, base = pushers)

grp_panel |>
  inner_join(grp_base, by = c("hdi_tier", "group")) |>
  mutate(index = pushers / base) |>
  ggplot(aes(date, index, color = hdi_tier)) +
  geom_hline(yintercept = 1, color = "grey70", linewidth = 0.3) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ group, scales = "free_y", ncol = 4) +
  scale_color_brewer(palette = "RdYlBu", name = "HDI tier") +
  labs(
    title    = "Family-level growth by HDI tier (indexed to 2020-Q1 = 1)",
    x = NULL, y = "Index"
  ) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))
```

## Takeaways

-   **Python and JavaScript are universal**, but their *mix* differs: Python
    is a larger share of activity in lower-HDI economies, while JavaScript /
    TypeScript skew toward higher-HDI ones — visible in both the heatmap
    and the LOESS curves.
-   **Rust, Go, Swift, Kotlin** sit clearly to the right of the dashed
    global-mean line in the dot-plot — they are programmer languages of
    high-HDI economies. **PHP** sits clearly to the left.
-   The growth panels show the **fastest pusher growth happening in
    Medium-HDI economies** since 2020, on top of a smaller base. Very-High
    HDI growth is real but slower.
-   For specific languages, indexed growth in Low/Medium-HDI tiers
    outpaces Very-High for Python, TypeScript and Rust — i.e. emerging
    markets are catching up on the "modern stack" faster than they are on
    older ones.
