---
title: "GitHub Topics by Economic Development (HDI)"
format:
  html:
    embed-resources: true
    toc: true
    df-print: paged
    fig-width: 9
    fig-height: 6
editor:
  markdown:
    wrap: 80
---

## Idea

Same setup as the languages-by-HDI notebook, but using **repository topics**
(`topics.csv`) instead of programming languages. Topics are user-applied tags
like `react`, `machine-learning`, `kubernetes`, `blockchain` — closer to what
people are *building* than what they're typing.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(scales)
library(countrycode)

theme_set(theme_minimal(base_size = 11))
```

## Data

```{r}
topics <- read_csv("topics.csv", show_col_types = FALSE)
hdi_raw <- read_csv("hdi.csv", show_col_types = FALSE)

hdi <- hdi_raw |>
  transmute(
    iso2_code = countrycode(iso3, "iso3c", "iso2c"),
    country   = country,
    region    = region,
    hdi_tier  = factor(hdicode, levels = c("Low", "Medium", "High", "Very High")),
    hdi       = hdi_2022
  ) |>
  filter(!is.na(iso2_code), !is.na(hdi))
```

The topics file includes an `EU` aggregate row — drop it so we don't double
count when joining to per-country HDI.

```{r}
topics <- topics |> filter(iso2_code != "EU")
```

Two extremely high-volume topics, `config` and `github-config`, are auto-tags
generated by GitHub default templates rather than human signal. Drop them.

```{r}
NOISE <- c("config", "github-config")
topics <- topics |> filter(!topic %in% NOISE)

glimpse(topics)
```

## Snapshot quarter

```{r}
latest <- topics |> slice_max(year * 10 + quarter, n = 1) |>
  distinct(year, quarter)
latest

snap <- topics |>
  semi_join(latest, by = c("year", "quarter")) |>
  inner_join(hdi, by = "iso2_code")
```

## Why low-HDI tiers look thin: data coverage by tier

**Important caveat.** The Innovation Graph uses a 100-developer privacy
floor: *"we decline to publish specific statistics for economies with fewer
than 100 unique developers performing the relevant activity during the
specified quarter."* So a country-topic-quarter cell with 99 developers
disappears entirely. Low-HDI economies hit this floor constantly,
especially on niche topics — the thinness of their rows below is the
**suppression rule**, not GitHub picking only "top repos".

The two unconditional totals — `developers.csv` and `repositories.csv` —
aren't sliced by topic, so they expose the real footprint of each tier.

```{r}
devs  <- read_csv("developers.csv",   show_col_types = FALSE) |>
  filter(iso2_code != "EU")
repos <- read_csv("repositories.csv", show_col_types = FALSE) |>
  filter(iso2_code != "EU")

devs_snap  <- devs  |> semi_join(latest, by = c("year", "quarter")) |>
  inner_join(hdi, by = "iso2_code")
repos_snap <- repos |> semi_join(latest, by = c("year", "quarter")) |>
  inner_join(hdi, by = "iso2_code")
```

### Country presence: HDI list vs. data

For each tier we count the countries on UNDP's HDI list, then how many of
them actually surface in the GitHub developers / topics tables for the
latest quarter.

```{r}
hdi_count <- hdi |> count(hdi_tier, name = "in_hdi")

dev_count <- devs_snap |> distinct(hdi_tier, iso2_code) |>
  count(hdi_tier, name = "in_devs")

topic_count <- snap |> distinct(hdi_tier, iso2_code) |>
  count(hdi_tier, name = "in_topics")

coverage <- hdi_count |>
  left_join(dev_count,   by = "hdi_tier") |>
  left_join(topic_count, by = "hdi_tier") |>
  mutate(across(c(in_devs, in_topics), \(x) replace_na(x, 0)))

coverage
```

```{r fig-coverage, fig.width=8, fig.height=4}
coverage |>
  pivot_longer(c(in_hdi, in_devs, in_topics),
               names_to = "set", values_to = "n") |>
  mutate(set = factor(set,
                      levels = c("in_hdi", "in_devs", "in_topics"),
                      labels = c("On UNDP HDI list",
                                 "Has any GitHub developers data",
                                 "Has any topics data"))) |>
  ggplot(aes(hdi_tier, n, fill = set)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = n), position = position_dodge(width = 0.9),
            vjust = -0.3, size = 3) +
  scale_fill_brewer(palette = "Blues", name = NULL) +
  labs(
    title    = "Country presence by HDI tier (latest quarter)",
    subtitle = "Topics data is sparser than developers data because of the 100-dev threshold per topic",
    x = "HDI tier", y = "Number of countries"
  ) +
  theme(legend.position = "bottom")
```

### Volume per tier

Total developers, repositories, and topic-tagged pushers in each tier.

```{r}
volume <- bind_rows(
  devs_snap  |> group_by(hdi_tier) |>
    summarise(value = sum(developers),   .groups = "drop") |>
    mutate(metric = "Developers"),
  repos_snap |> group_by(hdi_tier) |>
    summarise(value = sum(repositories), .groups = "drop") |>
    mutate(metric = "Repositories"),
  snap       |> group_by(hdi_tier) |>
    summarise(value = sum(num_pushers),  .groups = "drop") |>
    mutate(metric = "Topic-tagged pushers")
) |>
  mutate(metric = factor(metric,
                         levels = c("Developers", "Repositories",
                                    "Topic-tagged pushers")))

volume |>
  pivot_wider(names_from = metric, values_from = value) |>
  mutate(across(where(is.numeric), comma))
```

```{r fig-volume, fig.width=10, fig.height=4}
ggplot(volume, aes(hdi_tier, value, fill = hdi_tier)) +
  geom_col() +
  geom_text(aes(label = label_number(scale_cut = cut_short_scale())(value)),
            vjust = -0.3, size = 3) +
  facet_wrap(~ metric, scales = "free_y") +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale()),
                     expand = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "RdYlBu", guide = "none") +
  labs(
    title    = "Footprint per HDI tier (latest quarter)",
    x = "HDI tier", y = NULL
  )
```

### How many topics survive the privacy floor in each tier

```{r fig-topic-survival, fig.width=8, fig.height=4}
surv <- snap |> distinct(hdi_tier, topic) |>
  count(hdi_tier, name = "topics_with_data")

ggplot(surv, aes(hdi_tier, topics_with_data, fill = hdi_tier)) +
  geom_col() +
  geom_text(aes(label = comma(topics_with_data)), vjust = -0.3, size = 3) +
  scale_fill_brewer(palette = "RdYlBu", guide = "none") +
  scale_y_continuous(labels = label_comma(),
                     expand = expansion(mult = c(0, 0.1))) +
  labs(
    title    = "Distinct topics with at least one country reporting data",
    subtitle = "Lower-HDI tiers see far fewer topics survive the 100-developer floor",
    x = "HDI tier", y = "Distinct topics"
  )
```

### Per-country medians

Tier totals are dominated by the US, India, China, etc. Looking at the
median country in each tier separates "tier is small" from "tier is full
of small countries".

```{r}
devs_snap |>
  group_by(hdi_tier) |>
  summarise(
    countries        = n(),
    median_devs      = median(developers),
    `90th percentile` = quantile(developers, 0.9),
    max_devs         = max(developers),
    .groups = "drop"
  ) |>
  mutate(across(starts_with("median") | starts_with("90") | starts_with("max"),
                comma))
```

**Reading the table.** The Low tier's median country has only a few
thousand developers, well below what's needed to clear the 100-dev floor
on most individual topics. The Very-High tier's median is an order of
magnitude higher. The downstream charts in this notebook should be read
with that in mind — they show the *shape* of activity in each tier, not a
faithful sample of every economy in it.

## A curated focus set of topics

Themes most useful to compare across HDI:

```{r}
focus_topics <- c(
  # Web frontend
  "react", "vue", "angular", "nextjs", "tailwindcss",
  # Backend
  "nodejs", "django", "flask", "fastapi", "spring-boot", "laravel",
  # Mobile
  "android", "ios", "flutter", "react-native",
  # Data / ML / AI
  "machine-learning", "deep-learning", "ai", "llm",
  "tensorflow", "pytorch",
  # Infra / DevOps
  "docker", "kubernetes", "terraform", "aws",
  # Crypto / web3
  "blockchain", "ethereum", "solidity", "web3",
  # Security
  "security", "cybersecurity",
  # Community
  "hacktoberfest"
)

topic_theme <- tibble(
  topic = focus_topics,
  theme = c(
    rep("Web frontend",  5),
    rep("Backend",       6),
    rep("Mobile",        4),
    rep("Data/ML/AI",    6),
    rep("Infra/DevOps",  4),
    rep("Crypto/web3",   4),
    rep("Security",      2),
    rep("Community",     1)
  )
)
```

## Top topics globally (latest quarter)

```{r fig-global-top, fig.height=7}
snap |>
  group_by(topic) |>
  summarise(pushers = sum(num_pushers), .groups = "drop") |>
  slice_max(pushers, n = 25) |>
  ggplot(aes(pushers, fct_reorder(topic, pushers))) +
  geom_col(fill = "#2c7fb8") +
  scale_x_continuous(labels = label_comma()) +
  labs(
    title    = sprintf("Top 25 GitHub topics (%dQ%d)",
                       latest$year, latest$quarter),
    subtitle = "Pushers in repos tagged with each topic, summed across countries",
    x = "Pushers", y = NULL
  )
```

## Topic share within each HDI tier

```{r fig-tier-heatmap, fig.width=9, fig.height=10}
tier_totals <- snap |>
  group_by(hdi_tier) |>
  summarise(tier_total = sum(num_pushers), .groups = "drop")

tier_topic <- snap |>
  filter(topic %in% focus_topics) |>
  group_by(hdi_tier, topic) |>
  summarise(pushers = sum(num_pushers), .groups = "drop") |>
  inner_join(tier_totals, by = "hdi_tier") |>
  mutate(share = pushers / tier_total) |>
  inner_join(topic_theme, by = "topic") |>
  mutate(text_color = if_else(share > 0.5 * max(share), "white", "grey15"))

ggplot(tier_topic,
       aes(hdi_tier, fct_reorder(topic, share, .fun = max), fill = share)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = percent(share, accuracy = 0.01), color = text_color),
            size = 2.6) +
  scale_color_identity() +
  scale_fill_viridis_c(option = "mako", direction = -1, end = 0.85,
                       labels = label_percent(),
                       name = "Share of\ntier's pushers") +
  facet_grid(theme ~ ., scales = "free_y", space = "free_y") +
  labs(
    title    = "Topic share within each HDI tier",
    subtitle = "Each cell = % of pushers in that tier whose repos carry the topic",
    x = "HDI tier", y = NULL
  ) +
  theme(strip.text.y = element_text(angle = 0, face = "bold"),
        strip.background.y = element_rect(fill = "grey95", color = NA))
```

## HDI vs topic share, country by country

Every dot is a country; size = total pushers; line = pusher-weighted LOESS.
The *slope* of the line is what's interesting — sloping up means the topic is
disproportionately a high-HDI activity, sloping down means the opposite.

```{r fig-scatter, fig.width=11, fig.height=12}
country_totals <- snap |>
  group_by(iso2_code, hdi, hdi_tier) |>
  summarise(total = sum(num_pushers), .groups = "drop")

scat <- snap |>
  filter(topic %in% focus_topics) |>
  group_by(iso2_code, topic) |>
  summarise(pushers = sum(num_pushers), .groups = "drop") |>
  inner_join(country_totals, by = "iso2_code") |>
  mutate(share = pushers / total) |>
  filter(total >= 500) |>
  inner_join(topic_theme, by = "topic")

ggplot(scat, aes(hdi, share)) +
  geom_point(aes(size = total, color = hdi_tier), alpha = 0.55) +
  geom_smooth(aes(weight = total), method = "loess", se = FALSE,
              color = "grey25", linewidth = 0.5, formula = y ~ x) +
  facet_wrap(~ topic, scales = "free_y", ncol = 4) +
  scale_size_area(max_size = 5, labels = label_comma(),
                  name = "Pushers") +
  scale_y_continuous(labels = label_percent()) +
  scale_color_brewer(palette = "RdYlBu", name = "HDI tier") +
  labs(
    title    = "Topic share vs. HDI",
    subtitle = "One point per country (≥500 pushers); line = weighted LOESS",
    x = "HDI (2022)", y = "Share of country's pushers"
  ) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))
```

## Where each topic sits on the development gradient

Pusher-weighted mean HDI of each topic's contributors, grouped by theme.

```{r fig-mean-hdi, fig.width=9, fig.height=8}
topic_meanhdi <- snap |>
  filter(topic %in% focus_topics) |>
  group_by(topic) |>
  summarise(
    mean_hdi = weighted.mean(hdi, num_pushers),
    pushers  = sum(num_pushers),
    .groups  = "drop"
  ) |>
  inner_join(topic_theme, by = "topic")

global_mean <- weighted.mean(snap$hdi, snap$num_pushers)

ggplot(topic_meanhdi,
       aes(mean_hdi, fct_reorder(topic, mean_hdi),
           color = theme)) +
  geom_vline(xintercept = global_mean, linetype = "dashed",
             color = "grey50") +
  geom_segment(aes(x = global_mean, xend = mean_hdi, yend = topic),
               linewidth = 0.6) +
  geom_point(size = 3) +
  scale_color_brewer(palette = "Dark2", name = NULL) +
  annotate("text", x = global_mean, y = 1,
           label = sprintf("Global mean (%.3f)", global_mean),
           hjust = -0.05, vjust = 0, size = 3, color = "grey40") +
  labs(
    title    = "Pusher-weighted mean HDI of each topic's contributors",
    subtitle = "Right of dashed line = skews high-HDI; left = skews low-HDI",
    x = "Mean HDI", y = NULL
  ) +
  theme(legend.position = "bottom")
```

## Trends within tiers, 2020 → today

For each tier and topic, index quarterly pusher counts to the topic's first
non-zero quarter in that tier. Shows where each tier is *adopting* a topic
fastest.

```{r fig-trend-tier, fig.width=11, fig.height=10}
panel <- topics |>
  inner_join(hdi |> select(iso2_code, hdi_tier), by = "iso2_code") |>
  filter(topic %in% focus_topics) |>
  group_by(year, quarter, hdi_tier, topic) |>
  summarise(pushers = sum(num_pushers), .groups = "drop") |>
  mutate(date = yq(paste0(year, " Q", quarter)))

baseline <- panel |>
  filter(year == 2020, quarter == 1, pushers > 0) |>
  select(hdi_tier, topic, base = pushers)

panel <- panel |>
  inner_join(baseline, by = c("hdi_tier", "topic")) |>
  mutate(index = pushers / base)

ggplot(panel, aes(date, index, color = hdi_tier)) +
  geom_hline(yintercept = 1, color = "grey70", linewidth = 0.3) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~ topic, scales = "free_y", ncol = 4) +
  scale_color_brewer(palette = "RdYlBu", name = "HDI tier") +
  labs(
    title    = "Topic adoption growth by HDI tier (indexed to 2020-Q1 = 1)",
    subtitle = "Higher line = faster growth in that tier on that topic",
    x = NULL, y = "Index"
  ) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))
```

## Takeaways

-   **DevOps/Infra** topics (`kubernetes`, `terraform`, `aws`, `docker`) sit
    firmly on the right of the dashed line: they are
    high-HDI-economy activities, dominated by Very-High and High tiers.
-   **Crypto/web3** (`blockchain`, `ethereum`, `solidity`, `web3`) and
    **`hacktoberfest`** lean *the opposite way* — they over-index in lower
    HDI tiers relative to those tiers' total developer footprint.
-   **Mobile** splits cleanly: `ios` is a high-HDI topic, while `android`
    and `flutter` skew toward lower HDI — matching what we'd expect from
    device-affordability patterns.
-   **AI / LLM** topics (`ai`, `llm`, `deep-learning`) are growing
    everywhere, but the index plot shows the steepest *relative* growth
    in Medium-HDI economies — a catch-up effect on the newest stack.
-   Web-frontend topics (`react`, `nextjs`, `tailwindcss`) are spread
    fairly evenly, with `react` slightly higher-HDI and `tailwindcss`
    growing fastest in Medium-HDI tiers.
